---
title: "Bellabeat Project Report"
author: "Michalis Loizos"
date: "2025-04-22"
output:
  html_document:
    df_print: paged
latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(dplyr)
library(kableExtra)
library(RColorBrewer)
library(ggplot2)
library(tidyverse)
```



## Introduction
This project is my capstone for the Google Data Analytics Certificate, focusing on Bellabeat, a health tech company offering wellness products like the Leaf tracker. I analyzed Fitbit user data from 34 users, including daily activity (steps, calories), hourly trends, and workout intensities, to uncover patterns and provide marketing recommendations for Bellabeat’s target audience. Using Google Sheets for data cleaning and pivot table creation, SQL for filtering, and R (tidyverse, ggplot2) for visualizations, this project demonstrates skills in data wrangling, visualization, and business analysis.






## Data Processing and Summary
- This report summarizes the work performed on the Bellabeat dataset. The project used Google Sheets for data cleaning and pivot table creation, SQL to apply filters, and R for visualizations and the summary report.

- Data files from the Bellabeat dataset were carefully inspected for inconsistencies and cleaned as needed, such as converting UTC-formatted times to 24-hour format for hourly user trend studies. Afterward, a pivot table was created in Google Sheets to summarize user application data and identify the most active users. Table 1 shows the pivot table of user categories by calories burned

```{r}


pivot_table <- read.csv("PIVOT_TABLE.csv")

pivot_table_sorted <- pivot_table %>%
  arrange(desc(Average_Calories))

# Specify Columns
colnames(pivot_table_sorted) <- c("ID", 
                                  "Average Total Steps", 
                                  "Average Total Distance", 
                                  "Average Tracker Distance", 
                                  "Average Logged Activities Distance",
                                  "Average Calories",
                                  "Average Lightly Active Minutes",
                                  "Average Fairly Active Minutes",
                                  "Average Very Active Minutes",
                                  "Average Sedentary Minutes")

# Specify Column Format
kable(pivot_table_sorted, format = "latex", booktabs = TRUE,
      caption = "Pivot Table Showing User Categories by Calories Burned") %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8) %>%
  column_spec(1, width = "1.5cm") %>%  # Adjust width for ID column
  column_spec(2, width = "2cm") %>%    # Adjust width for Steps column
  column_spec(3, width = "2cm") %>%    # Adjust width for Total Distance column
  column_spec(4, width = "2cm") %>%    # Adjust width for Tracker Distance column
  column_spec(5, width = "2cm") %>% # Wider width for the longest column name
  column_spec(6, width = "2cm") %>%   
  column_spec(7, width = "2cm") %>%
  column_spec(8, width = "2cm") %>%
  column_spec(9, width = "2cm") %>%
  column_spec(10, width = "2cm")


```

## Most Active Users

To determine the most active users, the pivot table was sorted in descending order by average calories burned. All 34 users in the dataset were then categorized as "Lightly Active" (overall average calories burned: 0–2000), "Fairly Active" (2000–3000 calories), or "Very Active" (>3000 calories) over a 30-day period. The following SQL query was used to retrieve the data (commented out ):



```{r}
#SELECT
#ID,
#Average_Calories,

#CASE
#  WHEN Average_Calories BETWEEN 0 AND 2000 THEN 'Lightly Active'
#  WHEN Average_Calories BETWEEN 2000 AND 3000 THEN 'Fairly Active'
#  WHEN Average_Calories > 3000 THEN 'Very Active'

#END AS activity_level


#FROM 
#[Project_Name].Pivot_Table


#ORDER BY activity_level


```

The pie chart below shows the activity trend based on the overall average Calories.



```{r, fig.cap="User activity level categorization based on calories burned"}

data<- read.csv("CATEGORIZED_USERS-ON_ACTIVITY_LEVEL.csv")

library(ggplot2)
library(dplyr)


# Culculate the percentage for each activity level category
summary_data <- data %>%
  count(activity_level) %>%
  mutate(
    perc = n / sum(n) * 100,
    label = paste0(activity_level, "\n", round(perc, 0), "%")
  )

# Generate the Pie Chart
ggplot(summary_data, aes(x = "", y = n, fill = activity_level)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  geom_text(aes(label = label), position = position_stack(vjust = 0.5)) +
  theme_void() +
  labs(title = "User Activity Level Distribution") +
  scale_fill_brewer(palette = "RdBu")


```

It is evident from the pie chart that almost 50% of users are fairly active. The overall average calories were chosen as the metric to determine the most active users of the application, as this encompasses all types of activities. Using another metric like Logged Activities Distance would only include specific workouts and record activity initiated by the user.


## Hourly Trends

Next, hourly trends were analyzed to study the application’s 10 most active users and suggest improvements. The analysis began by tracking hourly trends based on user calories, performed separately for daytime and nighttime hours. The query below was used to retrieve user information in conjunction with the constructed pivot table (commented out)



```{r}

#SELECT
#  a.Id,
#  FORMAT_TIME("%H:%M:%S", TIME(a.ActivityHour)) AS Hour_Of_Day,
#  AVG(a.Calories) AS AverageCalories
#FROM 
#  `[Project_Name].Hourly_Calories` a
#INNER JOIN (
#  SELECT ID
#  FROM `[Project_Name].Pivot_Table`
#  WHERE Average_Calories >= 2000
#  ORDER BY Average_Calories DESC 
#  LIMIT 10
#) top_users
#ON a.Id = CAST(top_users.ID AS INT64)
#WHERE EXTRACT(HOUR FROM ActivityHour) <  12 (or > 12 for nighttime users)
#GROUP BY Id, Hour_Of_Day
#ORDER BY Id, Hour_Of_Day;


```

The average hourly calories burned by the top 10 users were then plotted for both daytime and nighttime:

```{r, fig.cap = "Daily calorie burning trends of the most active users during daytime." }


every_nth <- function(n) {
  function(x) x[seq(1, length(x), by = n)]
}


data2 <- read.csv("HOURLY_CALORIES_MOST_ACTIVE_USERS_DAY_TO_NIGHT.csv")
ggplot(data = data2) +
  
  geom_col(mapping = aes(x = Hour_Of_Day, y = AverageCalories, fill = AverageCalories), 
           width = 0.5) +
  facet_wrap(~Id) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Activity Hour", y = "Calories Burned", 
       title = "Hourly Calories Burned by Most Active Users-Daytime") +
  scale_fill_viridis_c() +
  scale_x_discrete(breaks = every_nth(2))




```



```{r, fig.cap = "Daily calorie burning trends of the most active users during nighttime."}


every_nth <- function(n) {
  function(x) x[seq(1, length(x), by = n)]
}


data2 <- read.csv("HOURLY_CALORIES_MOST_ACTIVE_USERS_NIGHT_TO_DAY.csv")
ggplot(data = data2) +
  
  geom_col(mapping = aes(x = Hour_Of_Day, y = AverageCalories, fill = AverageCalories), 
           width = 0.5) +
  facet_wrap(~Id) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Activity Hour", y = "Calories Burned", 
       title = "Hourly Calories Burned by Most Active Users-Nighttime") +
  scale_fill_viridis_c() +
  scale_x_discrete(breaks = every_nth(2))




```



It is evident that during daytime, 40% of the most active users exhibit a peak in their daily calorie burn. This peak could be due to exercise; however, these people could work nightshifts, so more metrics must be considered. As evident from Figure 2, 30% of users exhibited a peak during nighttime.


```{r, fig.cap="Evolution of average steps per hour during daytime for the most active     users"}

data3 <- read.csv("HOURLY_STEPS_MOST_ACTIVE_USERS_MORNING_TO_NIGHT.csv")

data3_avg <- data3 %>%
  group_by(Id,  Hour_Of_Day) %>%
  summarise(
    Avg_Steps = mean(Total_Steps, na.rm = TRUE)  # Average across days
  )


ggplot(data = data3_avg) +
  
  geom_col(mapping = aes(x = Hour_Of_Day, y = Avg_Steps, fill = Hour_Of_Day), 
           width = 0.5) +
  facet_wrap(~Id) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Activity Hour", y = "Steps", 
       title = "Hourly Steps  by Most Active Users-Daytime") +
  scale_x_discrete(breaks = every_nth(2))+
  scale_fill_viridis_d() 
  
```

```{r, fig.cap = "Average hourly steps of the most active users during nighttime"}

data10 <- read.csv("HOURLY_STEPS_MOST_ACTIVE_USERS_NIGHT_TO_MORNING.csv")

data10_avg <- data10 %>%
  group_by(Id,  Hour_Of_Day) %>%
  summarise(
    Avg_Steps = mean(Total_Steps, na.rm = TRUE)  # Average across days
  )


ggplot(data = data10_avg) +
  
  geom_col(mapping = aes(x = Hour_Of_Day, y = Avg_Steps, fill = Hour_Of_Day), 
           width = 0.5) +
  facet_wrap(~Id) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Activity Hour", y = "Steps", 
       title = "Hourly Steps  by Most Active Users-Nighttime") +
  scale_x_discrete(breaks = every_nth(2))+
  scale_fill_viridis_d() 

```

Figures 4 and 5 show the hourly evolution of average steps. The same 50% of users that exhibited peaks in their calorie burning also exhibit a peak in their hourly steps from late evening to night. These results demonstrate that these users are more active during this time period; however, it is still not clear if this is due to simple movement, activity, or workouts. I investigated this by analyzing workout intensity data.


























```{r, fig.cap = "Average workout intensities of the most active users during daytime"}


data11 <- read.csv("HOURLY_INTENSITIES_MOST_ACTIVE_USERS_DAY_TO_NIGHT.csv")

data11_avg <- data11 %>%
  mutate(Average_Intensity = as.numeric(as.character(Average_Intensity))) %>%  # Ensure numeric
  group_by(Id, Hour_Of_Day) %>%
  summarise(
    Avg_Intensity = mean(Average_Intensity, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(data = data11_avg) +
  geom_col(mapping = aes(x = Hour_Of_Day, y = Avg_Intensity, fill = Hour_Of_Day),
           width = 0.5) +
  facet_wrap(~Id) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Activity Hour", y = "Intensity", 
       title = "Hourly Intensities by Most Active Users – Daytime") +
  scale_x_discrete(breaks = every_nth(2))+
  scale_fill_viridis_d()

```
```{r, fig.cap = "Average workout intensities of most active users during nighttime."}


data12 <- read.csv("HOURLY_INTENSITIES_MOST_ACTIVE_USERS_NIGHT_TO_DAY.csv")

data12_avg <- data12 %>%
  mutate(Average_Intensity = as.numeric(as.character(Average_Intensity))) %>%  # Ensure numeric
  group_by(Id, Hour_Of_Day) %>%
  summarise(
    Avg_Intensity = mean(Average_Intensity, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(data = data12_avg) +
  geom_col(mapping = aes(x = Hour_Of_Day, y = Avg_Intensity, fill = Hour_Of_Day),
           width = 0.5) +
  facet_wrap(~Id) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Activity Hour", y = "Intensity", 
       title = "Hourly Intensities by Most Active Users – Nighttime") +
  scale_x_discrete(breaks = every_nth(2))+
  scale_fill_viridis_d()

```

Figures 6 and 7 show the workout intensities. The same 50% of users that exhibited a peak during daytime in their calorie burning and steps also exhibit a peak in their workout intensity. This confirms that the previous analysis is correlated with workouts, not simple movement due to running errands or working nightshifts.

## Conclusions

In this study, Bellabeat application data were analyzed. Users were categorized based on their daily habits, specifically daily calorie burning. The analysis focused on the 10 most active users to reveal insights from the most committed users. It was found that 50% of the most active users choose to work out in the late evening to nighttime period. I recommend sending notifications to users during this period as a reminder to exercise. Additionally, the application could offer suggestions for users’ daily nutritional habits, adjusted based on their workout hours.Future work could explore heart rate data or longer timeframes to further enhance insights.
